---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# etrm: Energy Trading and Risk Management in R

<!-- badges: start -->
[![R-CMD-check](https://github.com/sleire/etrm/workflows/R-CMD-check/badge.svg)](https://github.com/sleire/etrm/actions)
[![Build status](https://ci.appveyor.com/api/projects/status/un202j6f1ga7f3uc?svg=true)](https://ci.appveyor.com/project/sleire/etrm)
[![Lifecycle: Stable](https://img.shields.io/badge/lifecycle-stable-green.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![License:MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

## Overview

`etrm` is an R package with tools for trading and financial risk management in energy markets. The package currently offer tools for two main activities:

1. Construction of forward market curves
2. Portfolio insurance trading strategies for energy price risk management

## Installation

The development version can be installed from [GitHub](https://github.com/sleire/etrm) with:

``` r
devtools::install_github("sleire/etrm")
```
## Examples of use

The following sections will provide examples using some of the synthetic data sets included in the package.

#### 1. The Maximum Smoothness Forward Curve
A typical characteristic of energy commodities such as electricity and natural gas is that delivery takes place over a period in time, not on a single date. Listed futures contracts cover standardized periods, such as "Week", "Month", "Quarter", "Season" or "Year". The forward curve is an essential tool for pricing non-standard OTC contracts having any settlement period. An example of such standard energy market contracts can be found in the package data set `powfutures130513`.


```{r example, echo=FALSE, , out.width = "70%"}
library(etrm)
powfutures130513
```

The function `msfc()` will create an instance of the S4 class `MSFC` with generic methods `plot()`, `summary()` and `show()`.
In addition to the arguments from the list of contracts, the user may also provide a prior function to the calculation. This is relevant for markets with strong seasonality, such as power markets. The default value is `prior = 0`, but the user can provide any vector expressing a belief regarding the market to be combined with the observed prices. In the example below we have used a simple seasonal prior from the package `powpriors130513` data set.

```{r msfc_fut_wpri, out.width = "90%"}
fwd_fut_wpri <- msfc(tdate = as.Date("2013-05-13"),          # trading date
                     include = powfutures130513$Include,     # vector with TRUE/FALSE, include contract?
                     contract = powfutures130513$Contract,   # vector with contract names
                     sdate = powfutures130513$Start,         # vector with contract start dates
                     edate = powfutures130513$End,           # vector with contract end dates
                     f = powfutures130513$Closing,           # vector with contract closing prices
                     prior = powpriors130513$mod.prior       # prior function
                     )

plot(fwd_fut_wpri, legend = "", title = "MSFC with prior for power futures 2013-05-13")
```


The forward curve is calculated with the function

$f(t) = \lambda(t) + \epsilon(t)$

where $\lambda(t)$ is the prior supplied by the user and $\epsilon(t)$ is an adjustment function taking the observed prices into account. The `msfc()` function finds the smoothest possible adjustment function by minimizing the mean squared value of a spline function, while ensuring that the average value of the curve $f(t)$ is equal to contract prices used in the calculation for the respective time intervals. The number of polynomials used in the spline along with `head(prior)` and computed prices based on the curve are available with the `summary()` method:
```{r msfc_show}
summary(fwd_fut_wpri)
```


The calculation without prior function, for comparison:
```{r msfc_fut_npri, out.width = "90%"}
fwd_fut_npri <- msfc(tdate = as.Date("2013-05-13"),         # trading date
                     include = powfutures130513$Include,    # vector with TRUE/FALSE, include contract?
                     contract = powfutures130513$Contract,  # vector with contract names
                     sdate = powfutures130513$Start,        # vector with contract start dates
                     edate = powfutures130513$End,          # vector with contract end dates
                     f = powfutures130513$Closing,          # vector with contract closing prices
                     prior = 0                              # no prior function
                     )

plot(fwd_fut_npri, legend = "", title = "MSFC excluding prior for power futures 2013-05-13")
```


The daily forward curve values can be found along with the prior function and contracts used in the calculation with the `show()` method.
An instance of `MSFC` is a rather rich object, and further details regarding the calculation, spline coefficients, etc. can be found in the slots:

```{r msfc_slots}
slotNames(fwd_fut_wpri)
```



#### 2. Portfolio Insurance Trading Strategies for Energy Price Risk Management

Futures trading strategies for price risk management, for commercial hedgers with long or short exposure. All models below aim to achieve a favorable unit price for the energy portfolio, while preventing it from breaching a pre defined cap (floor). 

The functions

- `cppi()` - Constant Proportion Portfolio Insurance   
- `dppi()` - Dynamic Proportion Portfolio Insurance   
- `obpi()` - Option Based Portfolio Insurance         
- `shpi()` - Step Hedge Portfolio Insurance            
- `slpi()` - Stop Loss Portfolio insurance             

implement alternative approaches to achieve this goal. They return S4 objects of type `CPPI`, `DPPI`, `OBPI`, `SHPI` and `SLPI` respectively, with methods `plot()`, `summary()` and `show()`.


In our example, we will consider the CAL-06 contract in the synthetic `powcal` data set, and start trading 500 days prior to the contract expiry. For the `OBPI`strategy presented below, the target price is calculated as an expected cap (floor) given by the option premium-adjusted strike price selected for the delta hedging scheme within a standard Black-76 option pricing framework. The default strike price is set at-the-money. The user may express a view regarding future market development by deviating from this level.

```{r powcal_06, echo=FALSE, out.width = "90%"}
day06 <- powcal$Date[!is.na(powcal$`CAL-06`)]
cal06 <- powcal$`CAL-06`[!is.na(powcal$`CAL-06`)]
dat06 <- data.frame(Date = day06, CAL06 = cal06)
dat06 <- tail(dat06, 500)
#head(dat06)
#ggplot2::ggplot(dat06, ggplot2::aes(x = Date, y = CAL06)) + ggplot2::geom_line(color = "#F8766D") + ggplot2::ylab("Price") + ggplot2::xlab("")
```


```{r cal06_obpi_long, out.width = "95%"}
cal06_obpi_b <- obpi(q = 30,               # volume 30 MW (buyer)
                     tdate = dat06$Date,   # vector with trading days until expiry
                     f = dat06$CAL06,      # vector with futures price
                     k = dat06$CAL06[1],   # default option strike price at-the-money
                     vol = 0.2,            # annualized volatility, for the Black-76 delta hedging
                     r = 0,                # default assumed risk free rate of interest
                     tdays = 250,          # assumed trading days per year
                     daysleft = 500,       # number of days to expiry
                     tcost = 0,            # transaction cost, percentage of daily quoted price
                     int = TRUE            # integer restriction, smallest transacted unit = 1
                   )

plot(cal06_obpi_b, legend = "bottom", title = "OBPI strategy buyer CAL-06")
```

The `summary()` method:

```{r cal06_obpi_b_sumary}
summary(cal06_obpi_b)
```

The `show()`method provide details regarding daily values for market price, transactions, exposed volume, futures contract position, the target price and the calculated portfolio price. Further details for a specific instance of a trading strategy can be found in the slots, see for example:

```{r cal06_obpi_b_slots}
slotNames(cal06_obpi_b)
```


The strategy CAL-06 OBPI strategy from a sellers point of view:

```{r cal_obpi_short, out.width = "95%"}
cal06_obpi_s <- obpi(q = - 30,             # volume -30 MW (seller)
                     tdate = dat06$Date,   # vector with trading days until expiry
                     f = dat06$CAL06,      # vector with futures price
                     k = dat06$CAL06[1],   # default option strike price at-the-money
                     vol = 0.2,            # annualized volatility, for the Black-76 delta hedging
                     r = 0,                # default assumed risk free rate of interest
                     tdays = 250,          # assumed trading days per year
                     daysleft = 500,       # number of days to expiry
                     tcost = 0,            # transaction cost, percentage of daily quoted price
                     int = TRUE            # integer restriction, smallest transacted unit = 1
                   )

plot(cal06_obpi_s, legend = "bottom", title = "OBPI strategy seller CAL-06")
```





